Automatic Parallelization and Transparent Fault Tolerance

### 2016-06-11 pubs/tfp16/slides/slides.tex
touchup slides

\begin{frame}
\frametitle{Scientific Computing in our Microcosm}
\only<1-9>{\textbf{Local evolution of scientific computing}}
\begin{itemize}
  \item<2-9> Serial Fortran programs
    \\\only<2>{\ \ Simple, imperative.}
  \item<3-9> C
    \\\only<3>{\ \ Similar. More pleasant syntax (Free form source code!)}
  \item<4-9> Vendor communication libraries
    \\\only<4>{\ \  Non-portable, but parallel}
  \item<5-9> MPI everywhere (inter- and intra-node)
    \\\only<5>{\ \ De facto standard emerges to solve portability}
  \item<6-9> C++
    \\\only<6>{\ \ Object oriented goodness (and more!)}
  \item<7-9> MPI+\{Pthreads, OpenMP, OpenCL, CUDA, etc\}
    \\\only<7>{\ \ We augment MPI now with other explicit parallel models}
  \item<8-9> Parallel runtimes, e.g., Cilk Plus, Intel Threading Building Blocks,
    Stanford's Legion, etc.
    \\\only<8>{\ \ Automated scheduling for irregularly shaped problems. Task-based}
\end{itemize}
\only<9>{%
\textbf{
Currently MPI+threading model dominates.}}

@@

\begin{frame}
\frametitle{In Search of Automatic Parallelization\\
\emph{or at least automatic scheduling}}
  \begin{figure}
    \begin{center}
      \includegraphics[width=0.95\linewidth,height=1in]%
        {figures/S3Dtaskgraph.png}~\footnote{Courtesy Stanford Legion project}
\\[2.0ex]  % vertical space
      S3D task dependency, combustion chemistry calculation
    \end{center}
  \end{figure}
Simplest interesting chemistry, task graph much larger with more complex reactants.
\vspace{0.1in}
Schedule by hand?
\end{frame}

@@

\begin{frame}
  \frametitle{Transparent Fault Tolerance?}
  \begin{columns}[t]
    \begin{column}{.48\linewidth}
      Checkpoint:
      \begin{enumerate}
      \item Processes synchronize at predetermined point;
      \item Stop the world;
      \item Dump global state;
      \item Resume computation.
      \end{enumerate}
    \end{column}
    \onslide<2->{%
    \begin{column}[T]{.48\linewidth}
      Restart:
      \begin{enumerate}
      \item Observe that application has hung/crashed;
      \item Identify last valid (complete) checkpoint image;
      \item Re-launch application specifying checkpoint image.
      \end{enumerate}
    \end{column}}
  \end{columns}
\vspace{0.1in}
\onslide<3>{%
Inherent problems:
\begin{itemize}
\item Idle processors during relatively slow writes burn tight energy budgets.
\item MTTF is shrinking and C/R isn't scaling.
\end{itemize}}
\end{frame}

@@

\onslide<3->{%
  \emph{But}, these are not computer scientists:
  \begin{itemize}
  \item Passing functions as arguments is familiar (since early Fortran).
  \item Implicit space leaks are confusing.
  \item Non-strict evaluation is largely irrelevant
  \item Unboxed primitives and arrays of the same (unboxed vector) are the primary data.
  \end{itemize}}
\end{frame}

@@

\begin{frame}
\frametitle{Project} 
A \emph{light-weight} implementation of a pure, higher-order, polymorphic,
functional language and runtime system with which we can experiment with
automatic parallelization strategies with varying degrees of language
strictness, and secondarily, with mechanisms for transparent fault tolerance.
\begin{enumerate}
\item<2-> Haskell to STG (or Core) via GHC \emph{(todo)};
  \\\only<2>{%
    \ \ STG and Core are higher-order polymorphic functional languages in their own right.}
\item<3-> STG to C \emph{(serial done, parallel in progress)};
  \\\only<3>{%
    \ \ Fairly faithful to STG papers}
\item<4-> Mini-Haskell front end (done).
  \\\only<4>{%
    \ \ Close to ML, syntatically}
\item<5-> Fault tolerance \emph{TBD};
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Reinventing the wheel?}  Why not use Manticore, MultiMLton, F\#,
multicore OCaml, SAC etc., \onslide<3->{or even GHC in toto?}
\begin{itemize}
  \item<2-> We have a small cadre of scientists already interested in Haskell/GHC.
  \item<3-> Very light weight:  easy to instrument, modify;
  \item<4-> Know exactly what is going on under the hood;
  \item<5-> There exist conduits for getting student interns up to speed inexpensively;
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile] % for verbatim
\frametitle{All the myriad ways in C}
\begin{itemize}
\item<1-> setjmp/longjmp;
\only<1>{Significant overhead and bookkeeping}
\item<2-> giant switch;
\only<2>{Ugly hack, not really linkable}
\item<3-> computed goto (gcc extension);
\only<3>{Limited to gcc}
\item<4-> trampoline, the simplest of which parameterless C functions return the
  address (function pointer) of where to go to next:
\begin{verbatim}
      while (1) f = (*f)();
\end{verbatim}
\only<4>{Our intial implementation used this approach}
\item<5-> implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
\only<5>{Fun, but out of scope}
\item<6-> generate assembly code, e.g., LLVM, directly.
\only<6>{Ambitious. Haskell LLVM bindings are unreliable}
\end{itemize}
\end{frame}

@@

  \onslide<4->\textbf{Questions?}
\end{frame}

>>>>

\begin{frame}
\frametitle{Scientific Computing in our Microcosm}
\only<1-9>{\textbf{Local evolution of scientific computing}}
\begin{itemize}
  \item<2-9> Serial Fortran programs
    \\\only<2>{\ \ Simple, imperative.}
  \item<3-9> C
    \\\only<3>{\ \ Similar. More pleasant syntax (Free form source code!)}
  \item<4-9> Vendor communication libraries
    \\\only<4>{\ \  Non-portable, but parallel}
  \item<5-9> MPI everywhere (inter- and intra-node)
    \\\only<5>{\ \ De facto standard emerges to solve portability}
  \item<6-9> C++
    \\\only<6>{\ \ Object oriented goodness (and more!)}
  \item<7-9> MPI+\{Pthreads, OpenMP, OpenCL, CUDA, etc\}
    \\\only<7>{\ \ We augment MPI now with other explicit parallel models}
  \item<8-9> Parallel runtimes, e.g., Cilk Plus, Intel Threading Building Blocks,
    Stanford's Legion, etc.
    \\\only<8>{\ \ Automated scheduling for irregularly shaped problems.}
\end{itemize}
\only<9>{%
\textbf{
Currently MPI+threading model dominates.}}

@@

\begin{frame}
\frametitle{In Search of Automatic Parallelization\\
\emph{or at least automatic scheduling}}
  \begin{figure}
    \begin{center}
      \includegraphics[width=0.95\linewidth,height=1in]%
        {figures/S3Dtaskgraph.png}~\footnote{Courtesy Stanford Legion project}
\\[2.0ex]  % vertical space
      S3D task dependency, combustion chemistry calculation
    \end{center}
  \end{figure}
  \onslide<2->{%
    Simplest interesting chemistry, task graph much larger with more complex reactants.
    \vspace{0.1in}
    Schedule by hand?}
  \\\onslide<3->{%
    \textbf{This comes for free in the pure functional world}}
\end{frame}

@@

\begin{frame}
  \frametitle{Transparent Fault Tolerance?}
  \begin{columns}[t]
    \begin{column}{.48\linewidth}
      Checkpoint:
      \begin{enumerate}
      \item Processes synchronize at predetermined point;
      \item Stop the world;
      \item Dump global state;
      \item Resume computation.
      \end{enumerate}
    \end{column}
    \onslide<2->{%
    \begin{column}[T]{.48\linewidth}
      Restart:
      \begin{enumerate}
      \item Observe that application has hung/crashed;
      \item Identify last valid (complete) checkpoint image;
      \item Re-launch application specifying checkpoint image.
      \end{enumerate}
    \end{column}}
  \end{columns}
\vspace{0.1in}
\onslide<3->{%
Inherent problems:
\begin{itemize}
\item Idle processors burn tight energy budgets.
\item MTTF is shrinking and C/R isn't scaling.
\end{itemize}}
\onslide<4->{%
  \textbf{With referential transparency, this is a much smaller problem}}
\end{frame}

@@

\onslide<3->{%
  \emph{But}, these are not computer scientists:
  \begin{itemize}
  \item Passing functions as arguments is familiar (since early Fortran).
  \item Implicit space leaks are confusing.
  \item Non-strict evaluation is largely irrelevant.
  \item Unboxed primitives and arrays of the same (unboxed vector) are the primary data.
  \end{itemize}}
\end{frame}

@@

\begin{frame}
\frametitle{Project} 
A \emph{light-weight} implementation of a pure, higher-order, polymorphic,
functional language and runtime system with which we can experiment with
automatic parallelization strategies with varying degrees of language
strictness, and secondarily, with mechanisms for transparent fault tolerance.
\begin{enumerate}
\item<2-> Haskell to STG (or Core) via GHC \emph{(todo)};
  \\\only<2>{%
    \ \ STG and Core are higher-order polymorphic functional languages in their own right.}
\item<3-> STG to C \emph{(serial done, parallel in progress)};
  \\\only<3>{%
    \ \ Fairly faithful to STG papers}
\item<4-> Mini-Haskell front end (done);
  \\\only<4>{%
    \ \ Close to ML, syntatically}
\item<5-> Fault tolerance \emph{TBD}.
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Reinventing the wheel?}  Why not use Manticore, MultiMLton, F\#,
multicore OCaml, SAC etc., \onslide<3->{or even GHC in toto?}
\begin{itemize}
  \item<2-> We have a small cadre of scientists already interested in Haskell/GHC;
  \item<3-> Very light weight:  easy to instrument, modify;
  \item<4-> Know exactly what is going on under the hood;
  \item<5-> There exist conduits for getting student interns up to speed inexpensively.
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile] % for verbatim
\frametitle{All the myriad ways in C}
\begin{itemize}
\item<1-> setjmp/longjmp;
\\\only<1>{\ \ Significant overhead and bookkeeping}
\item<2-> giant switch;
\\\only<2>{\ \ Ugly hack, not really linkable}
\item<3-> computed goto (gcc extension);
\\\only<3>{\ \ Limited to gcc}
\item<4-> trampoline, the simplest of which parameterless C functions return the
  address (function pointer) of where to go to next:
\begin{verbatim}
      while (1) f = (*f)();
\end{verbatim}
\only<4>{\ \ Our intial implementation used this approach}
\item<5-> implement an intermediate language, e.g., C-{}-/Cmm, where tail call is primitive;
\\\only<5>{\ \ Fun, but out of scope}
\item<6-> generate assembly code, e.g., LLVM, directly.
\\\only<6>{\ \ Ambitious. Haskell LLVM bindings are unreliable}
\end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Thank you}
  \textbf{Questions?}
\end{frame}
